Probing the internal structure of quasars using microlensing events

### 2016-03-13 paper1/discussion.tex
Changes

\section{Discussion}
[Mihai's contribution]\\
In the current paper we simulate and study the resulting microlensing lightcurves of geometric crescent-shaped sources 
and compare them with the microlensing lightcurves of other simple mathematically describable source profiles. 
In order to mimic the behaviour of the flux of light from the source in the proximity of a fold caustic we make use of the simple approximation described in equation (5). 
The equation would exhaustively describe the magnification map and offer a good universal approximation for the 
particular microlensing regime that we consider. 
Namely, the shape of the caustic boundary in the proximity of the source in the respective plane can be approximated with
 a line due to reason that the local radius of curvature of caustic is orders of magnitude greater than the half-light
 radius of the studied source. 
In particular cases in which the previously mentioned approximation loses its validity the impact on the quality of the 
lightcurves is not evenly distributed. The shape of the lightcurve will be maintained. 
The data points corresponding to the source position before and during the overlapping of the caustic will be affected by
 smaller errors than the data points corresponding to later times. \\

@@

The previously mentioned abstract parameters can be related to physical quantities specific to the central region of a quasar. Thus the luminous region would correspond to the bright accretion disc that surrounds the black hole. The later's gravity would cause a shadow in the bright region limited by the extent of the event horizon of the black hole. Therefore the radius of the bright disk would provide an estimate of the size of the accretion disk and the radius of the dark disc would provide and estimate of the Schwarzschild radius of the black hole. Moreover a value of the Schwarzschild radius 
can be used to estimate the mass of the black hole $M \approx \frac{\Delta t_{dark}  v_p c^2}{2G}$. In the previous 
expression the $\Delta t_{dark}$ denotes the period of overlap between the black hole shadow and fold caustic. 
$v_p$ denotes the component of the relative velocity of the source and fold which is perpendicular to the caustic.  
The respective velocity is an unknown, though it can be constrained on a case by case basis to an order of magnitude or even better. This would require the 
study of the dynamics of the stellar structure which contains the gravitational lens. A better estimate of the relative velocity would facilitate a better estimate of the effective non-rotating black hole mass associated to the black hole shadow.

Furthermore, a simulated image of M87 presented in \citep{2012MNRAS.421.1517D} (figure 9) has been microlensed. On the resulting lightcurve the instances corresponding to the start and end of the black hole shadow and caustic overlap were distinguishable as can be seen in figure 10.

@@

In case of a high quality lightcurve with insignificant noise and measurement errors, the parameters can be obtained 
by simply identifying the characteristic instances of time without making use of the actual values of the magnification
 map. If the effect of the errors and the noise distorts the magnification time function enough so that the 
characteristic epochs are not identifiable with the characteristic periods still visible, the boundaries of the periods 
can be roughly estimated. Furthermore, if direct estimates of the parameters cannot be obtained we propose the use of a strong
statistical tool such as Markov-Chain Monte Carlo. In Section 6  we have studied the possibility of identifying a crescent 
source and  the possibility of recovering the respective parameters. As magnification map we have used a complex numerical 
one generated with the microlensing code by \cite{1999A&A...346L...5W}. In addition we have added to the signal a Gaussian 
noise with a SNR of 1.6.  In the experiment we have considered all nine combinations of original source profiles and assumed fitting source models.
Effectively we have fitted using MCMC all three sources with all three assumed fitting models. The results of the experiment allowed us build a 
procedure for distinguishing the shape of the source assuming that one of the three models we have considered is a good approximation.  
The procedure would be useful for observers that endeavour to gain more information about an unresolved source for which they can study 
the microlensing lightcurve.  As a first step in the procedure, one should first attempt to use MCMC with a Gaussian model assumption. 
If either the value of the $\chi^2$ or the number of rejected datapoints is large then the next step is to change the assumed source model to 
an uniform disc and redo the fitting. Finally if the uniform disc assumption is rejected as well then the fitting should be done with a 
crescent source model assumption. If at each of the three steps the data rejects the model then the source cannot be approximated by any of the three models. 
Otherwise if at one of the steps the data does not reject the model then the respective model is a good approximation.

[Contributions from the authors]\\
P.S. provided the original idea and plan for the research project as well as multiple contributions to the analysis. M.T. simulated and studied the microlensing 
lightcurves for the different source profiles discussed for clean data and approximated fold magnification amplification. J.W. provided the training and the
numerical code used by M. R. to create the magnification map and corresponding lighcurves used in the MCMC analysis, analysis that was performed by 
I. M. in the last part of the presented work.

>>>>

\section{Discussion}

In the current paper we simulate and study the resulting microlensing lightcurves of geometric crescent-shaped sources 
and compare them with the microlensing lightcurves of other simple mathematically describable source profiles. 
In order to mimic the behaviour of the flux of light from the source in the proximity of a fold caustic we make use of the simple approximation described in equation (5). 
The equation would exhaustively describe the magnification map and offer a good universal approximation for the 
particular microlensing regime that we consider. 
Namely, the shape of the caustic boundary in the proximity of the source in the respective plane can be approximated with
 a line due to reason that the local radius of curvature of caustic is orders of magnitude greater than the half-light
 radius of the studied source. 
In particular cases in which the previously mentioned approximation loses its validity the impact on the quality of the 
lightcurves is not evenly distributed. The shape of the lightcurve will be maintained. 
The data points corresponding to the source position before and during the overlapping of the caustic will be affected by
 smaller errors than the data points corresponding to later times. \\

@@

The previously mentioned abstract parameters can be related to physical quantities specific to the central region of a quasar. Thus the luminous region would correspond to the bright accretion disc that surrounds the black hole. The later's gravity would cause a shadow in the bright region limited by the extent of the event horizon of the black hole. \textbf{Therefore the radius of the bright disk would provide an estimate of the size of the accretion disk and the radius of the dark disc would provide and estimate of the gravitationally magnified Schwarzschild radius of the black hole $R_{S}^{magnified} = \Delta t_{dark}*v_p$. By $R_{S}^{magnified}$ we refer to the apparent Schwarzschild radius which is larger than the real value at large distances due to the black hole's own gravity.  Moreover the gravitationally magnified value of the Schwarzschild radius is a monotonic function of the black hole's mass. Therefore it 
can be used to estimate the mass of the black hole if it was not rotating.} In the previous 
expression the $\Delta t_{dark}$ denotes the period of overlap between the black hole shadow and fold caustic. 
$v_p$ denotes the component of the relative velocity of the source and fold which is perpendicular to the caustic.  
The respective velocity is an unknown, though it can be constrained on a case by case basis to an order of magnitude or even better. This would require the 
study of the dynamics of the stellar structure which contains the gravitational lens. A better estimate of the relative velocity would facilitate a better estimate of the effective non-rotating black hole mass associated to the black hole shadow.

\textbf{In case of a high quality lightcurve with insignificant noise and measurement errors, the parameters can be obtained 
by simply identifying the characteristic instances of time without making use of the actual values of the magnification
 map. If the effect of the errors and the noise distorts the magnification time function enough so that the 
characteristic epochs are not identifiable with the characteristic periods still visible, the boundaries of the periods 
can be roughly estimated. Furthermore, if direct estimates of the parameters cannot be obtained we propose the use of a strong
statistical tool such as Markov-Chain Monte Carlo. In Section 6  we have studied the possibility of identifying a crescent 
source and  the possibility of recovering the respective parameters. As magnification map we have used a complex numerical 
one generated with the microlensing code by \cite{1999A&A...346L...5W}. In addition we have added to the signal a Gaussian 
noise with a SNR of 1.6.  In the experiment we have considered all nine combinations of original source profiles and assumed fitting source models.
Effectively we have fitted using MCMC all three sources with all three assumed fitting models. The results of the experiment allowed us build a 
procedure for distinguishing the shape of the source assuming that one of the three models we have considered is a good approximation.  
The procedure would be useful for observers that endeavour to gain more information about an unresolved source for which they can study 
the microlensing lightcurve.  As a first step in the procedure, one should first attempt to use MCMC with a Gaussian model assumption. 
If either the value of the $\chi^2$ or the number of rejected datapoints is large then the next step is to change the assumed source model to 
an uniform disc and redo the fitting. Finally if the uniform disc assumption is rejected as well then the fitting should be done with a 
crescent source model assumption. If at each of the three steps the data rejects the model then the source cannot be approximated by any of the three models. 
Otherwise if at one of the steps the data does not reject the model then the respective model is a good approximation.   }

\textbf{Furthermore, a simulated image of M87 presented in \citep{2012MNRAS.421.1517D} (figure 9) has been microlensed. On the resulting lightcurve the instances corresponding to the start and end of the black hole shadow and caustic overlap were distinguishable as can be seen in figure 10.}

@@

### 2016-03-13 paper1/intro.tex
Changes

Active galactic nucleii are thought to be powered through the
accretion of matter from the proximal environment into a supermassive
black hole.  The radiation emitted excites the surrounding medium
which becomes detectable as narrow line regions, broad line regions
and optical continuum.  Moreover, in the direction perpendicular to
the accretion disk, where the medium is more transparent, jets will
appear.  If a jet is oriented towards the Earth, a quasar is observed
\citep[e.g.,][]{1984RvMP...56..255B}.  While the basic mechanism
\citep[originating in the work
  of][]{1964ApJ...140..796S,1964SPhD....9..246Z,1969Natur.223..690L}
is not in doubt, the central engines, near the event horizons of the
black holes, remain to be probed.

For the black holes in the Galactic centre and the centre of M87
---the two nearest objects that barely qualify as AGN--- the central
engines (which are $<0.1\rm\,mas$ on the sky) are close to being
resolved through very long baseline interferometry, which shows
preliminary indications of the jet-launching structures
\citep{2008JPhCS.131a2055D,2012Sci...338..355D,2013MNRAS.434..765K,2016arXiv160205527F}.
Current data do not deliver images, but require fitting to to
predefined models for the images.  A whole range of models have been
applied, starting from simple geometric models to more complex
physical models
\citep{2008Natur.455...78D,2011ApJ...738...38B,2009ApJ...706..497M,2010ApJ...717.1092D}.
The more complicated models nonetheless tend to predict a crescent
shaped silhouette of the black hole.  This motivated
\cite{2013MNRAS.434..765K} to use a simple geometric crescent model to
fit the data, and argued that the crescent is nothing but the
silhouette of the event horizon.

@@

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, refers to sharp changes in the
observed brightness of quasars that have been lensed by intervening
galaxies, without any changes in the intrinsic luminosity.
Microlensing affects only the light from the innermost part of the
quasar, such as the optical continuum
\citep[e.g.,][]{2012A&A...544A..62S} and is a consequence of two
things: the very small size of the central engine, and granularity of
the mass distribution of a lensing galaxy due to stars.  The latter
means that the brightness of a lensed object (or lensing
magnification) is not a smooth function of source position, but a
contains a complicated network of singular curves, known as caustics.
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  The effect supplies an upper limit on the size of the
central engine, and can also be used to study the mass distribution
and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Caustics have an additional
remarkable feature: though they can be very complicated, they have
some universal properties well-known from catastrophe theory.  In
particular, very close to the simplest caustics (known as folds), the
magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

>>>>

Active galactic nuclei are thought to be powered through the
accretion of matter from the proximal environment into a supermassive
black hole.  The radiation emitted excites the surrounding medium
which becomes detectable as narrow line regions, broad line regions
and optical continuum.  Moreover, in the direction perpendicular to
the accretion disk, where the medium is more transparent, jets will
appear.  If a jet is oriented towards the Earth, a quasar is observed
\citep[e.g.,][]{1984RvMP...56..255B}.  While the basic mechanism
\citep[originating in the work
  of][]{1964ApJ...140..796S,1964SPhD....9..246Z,1969Natur.223..690L}
is not in doubt, the central engines, near the event horizons of the
black holes, remain to be probed.

For the black holes in the Galactic centre and the centre of M87
---the two nearest objects that barely qualify as AGN--- the central
engines (\textbf{which are $<0.1\rm\,mas$ on the sky}) are close to being
resolved through very long baseline interferometry, which shows
preliminary indications of the jet-launching structures
\citep{2008JPhCS.131a2055D,2012Sci...338..355D,2013MNRAS.434..765K,2016arXiv160205527F}.
Current data do not deliver images, but require fitting to
predefined models for the images.  A whole range of models have been
applied, starting from simple geometric models to more complex
physical models
\citep{2008Natur.455...78D,2011ApJ...738...38B,2009ApJ...706..497M,2010ApJ...717.1092D}.
The more complicated models nonetheless tend to predict a crescent
shaped silhouette of the black hole.  This motivated
\cite{2013MNRAS.434..765K} to use a simple geometric crescent model to
fit the data, and argued that the crescent is nothing but the
silhouette of the event horizon.

@@

Gravitational microlensing of quasars, reviewed in
Section~\ref{sec:microlensing} below, refers to sharp changes in the
observed brightness of quasars that have been lensed by intervening
galaxies, without any changes in the intrinsic luminosity.
Microlensing affects only the light from the innermost part of the
quasar, such as the optical continuum
\citep[e.g.,][]{2012A&A...544A..62S} and is a consequence of two
things: the very small size of the central engine, and granularity of
the mass distribution of a lensing galaxy due to stars.  \textbf{The latter
means that the local magnification is not a smooth function of source 
position. It contains a complicated network of singular curves, known as caustics.}
Figure~\ref{fig:magnification_map} shows part of a magnification map
with a few caustics.  The lensed brightness would be given by placing
the source on such a magnification map and integrating the surface
brightness weighted by the magnification.  Most astrophysical sources
straddle several caustics, and hence their net brightness varies
smoothly with location.  The central engine of quasars, however, is
smaller than the typical spacing between caustics.  As a result, the
lensed brightness undergoes sudden changes as a quasar crosses a
caustic.  The effect supplies an upper limit on the size of the
central engine, and can also be used to study the mass distribution
and kinematics of stars in the lensing galaxy as well
\citep[e.g.,][]{2012ApJ...744..111P}.  Caustics have an additional
remarkable feature: though they can be very complicated, they have
some universal properties well-known from catastrophe theory.  In
particular, very close to the simplest caustics (known as folds), the
magnification is approximately constant on one side and
$\propto1/\sqrt p$ where $p$ is the transverse distance of the source
from the caustic.  This property will be exploited later.

@@

\textbf{Finally in Section 7 we discuss in more depth the 
implications of the results presented in the previously mentioned sections.
One of the most interesting implication is the possibility to estimate the black hole
mass from the reconstructed parameters. This further requires good approximations of the 
relative transversal velocity between the quasar and the lens. Proper constraints can be set
with independent observations of the stellar structure that contains the gravitational lens.}

\textbf{The thorough study of the possibility to reconstruct the quasar's structural 
parameters from light-curves containing multiple microlensing events represents 
the target of future work. This will most probably require the use of powerful
statistical tools. Another path for future work is the designing of an observation regime 
best suited for acquiring the necessary microlensed light-curve .    }

### 2016-03-13 paper1/theory.tex
Changes

There are four characteristic points visible on the resulting one dimensional light profile of the crescent shaped source. Two of the points, p1 and p4, mark the outer boundaries of the luminous disc component. The other two, p2 and p3, mark the boundaries of the dark disc component. Since $S_1D$ is a projection of $S_2D$ on a line perpendicular to the caustic, the following relation holds:
\begin{equation}
	p_2-p_1 = R_p -R_n - a.
\end{equation}
In addition there are two other obvious relations the characteristic points which are independent of the projection:
\begin{equation}
	p_4 -p_1 = 2 R_p,
\end{equation}

@@

\begin{figure}
\centering
        \includegraphics[width = 0.48\textwidth]{figures/M87_image.eps}
\caption{\label{fig:M87_image} The simulated image of M87's black hole sillouette obtained by \citep{2012MNRAS.421.1517D} from a GRMHD simulation presented in \citep{2009MNRAS.394L.126M}. The color bar is linear and scaled to reach value 1 for the maximum brightness. Points p2 and p3, mark the start and end of the equivalent dark disk and are associated to the moment of their overlap with the caustic t2 and t3.}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{figures/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{figures/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained from the numerical integration along the ordonata axis of the source image (figure 9) is presented in the left subplot. The corresponding lightcurve associated to the 1D profile is presented in the right subplot. The points p2 and p3 mark the start and end of the equivalent dark disk and are associated to the moments of their overlap with the caustic: t2 and t3.}
\end{figure}

>>>>

There are four characteristic points visible on the resulting one dimensional light profile of the crescent shaped source. Two of the points, p1 and p4, mark the outer boundaries of the luminous disc component. The other two, p2 and p3, mark the boundaries of the dark disc component. Since $S_{1D}$ is a projection of $S_{2D}$ on a line perpendicular to the caustic, the following relation holds:
\begin{equation}
	p_2-p_1 = R_p -R_n - a.
\end{equation}
In addition there are two other obvious relations the characteristic points which are independent of the projection:
\begin{equation}
	p_4 -p_1 = 2 R_p,
\end{equation}

@@

\begin{figure}
\centering
        \includegraphics[width = 0.48\textwidth]{figures/M87_image.eps}
\caption{\label{fig:M87_image} The simulated image of M87's black hole
  sillouette obtained by \citep{2012MNRAS.421.1517D} from a GRMHD
  simulation presented in \citep{2009MNRAS.394L.126M}. The color bar
  is linear and scaled to reach value 1 for the maximum
  brightness.}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{figures/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{figures/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained
  from the numerical integration along the ordonata axis of the source
  image (figure 9) is presented in the lower panel. The corresponding
  lightcurve associated to the 1D profile is presented in the right
  panel. The points p2 and p3 mark the start and end of the equivalent
  dark disk and are associated to the moments of their overlap with
  the caustic: t2 and t3.}
\end{figure}
